Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-02 paper/main.tex
first part of section 6 updated

\title{Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving}
\author{Tereza Schwarzov\'{a}}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
\funding{T.~Schwarzová received funding from
the European Union’s Horizon Europe program under the Grant Agreement
No.\ 101087529.}

@@

\begin{abstract}
  This paper presents a novel application of QBF solving to automata
  reduction. We focus on \emph{Transition-based Emerson-Lei automata
    (TELA)}, which is a popular formalism that generalizes many
  traditional kinds of automata over infinite words including Büchi,
  co-Büchi, Rabin, Streett, and parity automata. Transitions in a TELA
  are labelled with acceptance marks and its accepting formula is a
  positive Boolean combination of atoms saying that a particular mark
  has to be visited infinitely or finitely
  often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique for reducing the
  number of %different
  acceptance marks in TELA based on satisfiability of \emph{quantified
    Boolean formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  %
  % In fact, we present three formula constructions
  % % trading speed for higher reduction efficiency.
  % with increasing reduction potential and formula complexity, and thus
  % also decreasing speed of their satisfiability solving.
  We evaluated
  our reduction technique on TELA produced by state-of-the-art tools
  of the libraries Owl and Spot and by the tool \texttt{ltl3tela}. The
  technique reduced some acceptance marks in automata produced by all
  the tools. On automata with more than one acceptance mark produced
  by tools Delag and Rabinizer~4 of the Owl library, our technique reduced
  29.1\% and 38.5\% of acceptance marks, respectively.
  % The achieved
  % reduction of acceptance marks on automata produced by tools of the Owl library was
  % between 25 and 40\%.
\end{abstract}

@@

We have implemented our reduction technique in a tool called
\telatko. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
Rabinizer~4~\cite{kretinsky.18.cav} (both included in the Owl
library), Spot~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}.  While the reduction is
relatively modest on TELA produced by \texttt{ltl3tela} and Spot, it
is substantial on automata produced by the tools of the Owl library.

@@

To evaluate our reduction technique, we considered all sets of LTL
formulas from literature as they are provided by the tool
\texttt{genltl} of the Spot library~\cite{duret.16.atva2}~2.10.4, where
parametrized formulas were generated for all combinations of
parameter values from 1 to 4. We used the tool \texttt{ltlfilt} of the
Spot library to simplify the formulas and remove
duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Three of the considered
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance marks as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators. For each translator, it also provides the number of
fails, the number of produced automata with less than two marks, and
the number of automata with at least two marks. The last number is
typeset in bold as these automata are actually used for the
experimental evaluation of our reduction technique.

\begin{table}[t]
  \caption{LTL to automata translators used for the experimental
    evaluation and the numbers of fails and successfully constructed
    automata with less than two marks and with at least two marks.}
\label{tab:ltltranslators}
\centering
% \begin{tabular}{lr@{~~~~~~}lr}
% \toprule
% tool        & version&homepage & ~~~automata \\
% \midrule
% \texttt{ltl2dela}~~~ & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 129\\
% \texttt{ltl2dgra} & 21.0 & & 234\\
% \texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 70\\
% \texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 91\\
% \bottomrule
% \end{tabular}
%
\begin{tabular}{lrrrr}
  \toprule
  &&& \multicolumn{2}{c}{constructed automata with} \\ \cmidrule(lr){4-5}
  tool &\!\!\!version/homepage & ~~~~fails & less than 2 marks & \textbf{at least 2 marks}\\
  \midrule
  \texttt{ltl2dela} & 21.0\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
       & 5 & 214 & \textbf{129} \\
  \texttt{ltl2dgra} & 21.0\footnotemark[1]
       & 12 & 102 & \textbf{234}\\
  \texttt{ltl2tgba} & 2.10.4\tablefootnote{\url{https://spot.lrde.epita.fr}}
       & 0 & 278 & \textbf{70}\\
  \texttt{ltl3tela} & 2.2.0\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
       & 18 & 239 & \textbf{91}\\
  \bottomrule
\end{tabular}
\end{table}

To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator), we have applied all
single-level reductions and the multi-level reduction, always with
incremental approach and without reducing the number of cubes. We
do not reduce the number of cubes as our primary aim is to reduce the
number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

>>>>

\title{Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving}
\author{Tereza Schwarzov\'{a}}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{https://orcid.org/0009-0001-9016-5399}{}
\author{Jan Strej\v{c}ek}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
\funding{T.~Schwarzová received funding from
the European Union’s Horizon Europe program under the Grant Agreement
No.\ 101087529.}

@@

\begin{abstract}
  This paper presents a novel application of QBF solving to automata
  reduction. We focus on \emph{Transition-based Emerson-Lei automata
    (TELA)}, which is a popular formalism that generalizes many
  traditional kinds of automata over infinite words including Büchi,
  co-Büchi, Rabin, Streett, and parity automata. Transitions in a TELA
  are labelled with acceptance marks and its accepting formula is a
  positive Boolean combination of atoms saying that a particular mark
  has to be visited infinitely or finitely
  often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique for reducing the
  number of %different
  acceptance marks in TELA based on satisfiability of \emph{quantified
    Boolean formulas (QBF)}.
  % : if a constructed QBF formula is satisfiable then
  % its model encodes an equivalent accepting formula with less
  % acceptance marks.
  %
  % In fact, we present three formula constructions
  % % trading speed for higher reduction efficiency.
  % with increasing reduction potential and formula complexity, and thus
  % also decreasing speed of their satisfiability solving.
  We evaluated
  our reduction technique on TELA produced by state-of-the-art tools
  of the libraries Owl and Spot and by the tool \texttt{ltl3tela}. The
  technique reduced some acceptance marks in automata produced by all
  the tools. \fixme{On automata with more than one acceptance mark produced
  by tools Delag and Rabinizer~4 of the Owl library, our technique reduced
  29.1\% and 38.5\% of acceptance marks, respectively.}
  % The achieved
  % reduction of acceptance marks on automata produced by tools of the Owl library was
  % between 25 and 40\%.
\end{abstract}

@@

We have implemented our reduction technique in a tool called
\telatko. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
Rabinizer~4~\cite{kretinsky.18.cav} (both included in the Owl
library), Spot~\cite{duret.16.atva2}, and
\texttt{ltl3tela}~\cite{major.19.atva}.  While the reduction is
relatively modest on TELA produced by \texttt{ltl3tela} and Spot, it
is substantial on automata produced by the tools of the Owl library.

@@

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 366 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G}) from the Spot library~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA
by \texttt{ltl3tela} with option \texttt{-D1} and by another two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}. Some translators failed on some
formulas: it usually reached a timeout of 60 seconds or it produced an
automaton that cannot be parsed by the Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of the
translators. For each set of formulas and for each translator, it also
provides the number of fails, the number of produced automata with
less than two marks, and the number of automata with at least two
marks. The numbers of automata with at least two marks are typeset in
bold as these automata are actually used for the experimental
evaluation of our reduction technique.

\begin{table}[t]
  \caption{Considered LTL to automata translators and the numbers of
    fails and successfully constructed automata with at most one mark and with at least two marks for each translator and set of
    formulas.}
\label{tab:ltltranslators}
\centering
% \begin{tabular}{lr@{~~~~~~}lr}
% \toprule
% tool        & version&homepage & ~~~automata \\
% \midrule
% \texttt{ltl2dela}~~~ & 21.0 & \multirow{2}{*}{\url{https://owl.model.in.tum.de/}} & 129\\
% \texttt{ltl2dgra} & 21.0 & & 234\\
% \texttt{ltl2tgba} & 2.10.4 & \url{https://spot.lrde.epita.fr} & 70\\
% \texttt{ltl3tela} & 2.2.0  & \url{https://github.com/jurajmajor/ltl3tela} & 91\\
% \bottomrule
% \end{tabular}
%
% \begin{tabular}{lrrrr}
%   \toprule
%   &&& \multicolumn{2}{c}{constructed automata with} \\ \cmidrule(lr){4-5}
%   tool &\!\!\!version/homepage & ~~~~fails & less than 2 marks & \textbf{at least 2 marks}\\
%   \midrule
%   \texttt{ltl2dela} & 21.0\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
%        & 5 & 214 & \textbf{129} \\
%   \texttt{ltl2dgra} & 21.0\footnotemark[1]
%        & 12 & 102 & \textbf{234}\\
%   \texttt{ltl2tgba} & 2.10.4\tablefootnote{\url{https://spot.lrde.epita.fr}}
%        & 0 & 278 & \textbf{70}\\
%   \texttt{ltl3tela} & 2.2.0\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
%        & 18 & 239 & \textbf{91}\\
%   \bottomrule
% \end{tabular}
%

\setlength{\tabcolsep}{5.7pt}
\begin{tabular}{lrrrrrrr}
  \toprule
  && \multicolumn{3}{c}{\!348 formulas from literature\!} & \multicolumn{3}{c}{366 random formulas}\\
  \cmidrule(lr){3-5} \cmidrule(lr){6-8}
  &&& \multicolumn{2}{c}{automata with} && \multicolumn{2}{c}{automata with}\\
  \cmidrule(lr){4-5} \cmidrule(lr){7-8}
  &&& at most & \textbf{at least} && at most & \textbf{at least}\\[-.8ex]
  \multicolumn{2}{l}{translator ~~~ (version)$^{\text{web}}$} & ~~~~fails & 1 mark & \!\!\!\!\textbf{2 marks} & ~~~~fails & 1 mark & \!\!\!\!\textbf{2 marks}\\
  \midrule 
  \texttt{ltl2tgba -G} & (2.10.4)\tablefootnote{\url{https://spot.lrde.epita.fr}}
  & 0 & 278 & \textbf{70} & ? & ? & \textbf{?}\\
  \texttt{ltl3tela} & (2.2.0)\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
  & 18 & 239 & \textbf{91} & ? & ? & \textbf{?}\\
  \texttt{ltl3tela -D1} & (2.2.0)\tablefootnote{\url{https://github.com/jurajmajor/ltl3tela}}
  & ? & ? & \textbf{?} & ? & ? & \textbf{?}\\ 
  \texttt{ltl2dela} & (21.0)\tablefootnote{\label{fn:owl}\url{https://owl.model.in.tum.de/}}
       & 5 & 214 & \textbf{129} & ? & ? & \textbf{?} \\
  \texttt{ltl2dgra} & (21.0)\footnotemark[4]
       & 12 & 102 & \textbf{234} & ? & ? & \textbf{?}\\
  \bottomrule
\end{tabular}
\end{table}

\fixme{Honza je tu.}
To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator), we have applied all
single-level reductions and the multi-level reduction, always with
incremental approach and without reducing the number of cubes. We
do not reduce the number of cubes as our primary aim is to reduce the
number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.
